# Getting Started {#GettingStarted}

## Packages

When you start R studio, basic functionality is initially available.  However, in most projects we will want to use some special code and functions contained in packages that are not initially available whe R starts.  Before packages can be used in our analyses, they must be installed in our R workspace.  We presume that the R Studio development environment is being used by our students.  Any package can be installed by clicking the "Packages" tab in the lower right panel of the R Studio workspace.  Then click "Install" to produce an entry bar where you type the name of the desired package.  

Or you can type a command to install a package:

```{r,eval=FALSE,echo=TRUE}
install.packages("alr4")
```


Once a package is installed into your R Studio environment, you make it available by loading with the `library()` command.  For this document, some additional packages are needed, and are loaded in the next code block.  The *knitr* and *tidyverse* packages have been previously installed.  If you attempt to load a package (in this case zelig) that has not been installed, you will get an error message similar to this:

`Error in library(zelig) : there is no package called ‘zelig’`


```{r , echo=TRUE,warning=FALSE,message=FALSE}
# this block loads R packages that may be needed for the analysis.
library(knitr)
library(tidyverse)
```


## The tidyverse Package

The tidyverse package is very special - it is a package of other packages. The tidyverse website [tidyverse](http://tidyverse.org "Title") describes the tidyverse as: *The tidyverse is an opinionated collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures.*

The most important packages inside the tidyverse package for this document are:  dplyr, magrittr, and ggplot2.


## Gapminder Data:

This dataset (named gapminder) is contained in an R package called *gapminder*, and needs to be loaded before the dataset can be used.

```{r}
library(gapminder)
```

## Set Working Directory

In the "Files" tab in the lower right portion of the R Studio work area, you can choose where you want to store files and conduct your work by navigating to a suitable folder by making folders and sub-folders and then clicking to navigate to a suitable work area.

You should notify R Studio and the R software to this location called the working directory.  Once you have navigated to where you want files, data, and results to reside, you notify R Studio by clicking the blue "More" gear and choose "set working directory."   This will help R understand where to expect files and dat to be located.

One of the most common problems students experience is that they work on files in a location not specified as the "working directory."

## Reading Data From a CSV File

The most common way to read data into R is from an excel spreadsheet that has been saved into a comma-separated-values (csv) file.  This means that data elements are separated from each other by commas ",".

We consider a data file named (file.csv) that contains variable names in the first row of the file. Place this file in your working directory and read,

```{r,eval=FALSE}
dataframe <- read.csv("file.csv",header=TRUE)
```

A frequent issue with `read.csv` is that character variables are automatically converted to factor/categorical variables.  This may not be a good choice in many instances.  To gain full control of how this is handled, you can prevent this kind of auto-conversion by using the `stringsAsFactors` option.

```{r,eval=FALSE}
cardata <- read.csv(file = 'carspeeds.csv', stringsAsFactors = FALSE)
```

The *readr* package inside the *tidyverse* family of packages has a slightly nicer read csv function you should know about.  We use the `readr::` prefix to inform readers that the `read_csv` function resides in the *readr* package.  This read function will not auto-convert character variables to category/factor variables.

```{r,eval=FALSE}
dataframe <- readr::read_csv("file.csv",col_names = TRUE)
```


Reading data directly from excel spreadsheets is more complex and you should read documentation for the *readxl* package.


