# Overview of a Dataframe {#OverviewDataframe}

Datasets in R are usually called dataframes or tibbles.  The distinction between these names is not important for our purposes - we will usually refer to a dataset as a dataframe.

```{r,echo=FALSE,message=FALSE,warning=FALSE}
library(gapminder)
library(tidyverse)
```


## glimpse 
Let's look at what is inside the gapminder dataset using the `glimpse` command from the *dplyr* package.  The *dplyr* package is contained in the package "tidyverse" that was loaded previously. The `glimpse(gapminder)` command would have executed without any errors. We use the `dplyr::` prefix to inform readers that the glimpse function resides in the *dplyr* package.

```{r}
# the next command would also execute if
# dplyr or tidyverse was loaded..
#glimpse(gapminder)
dplyr::glimpse(gapminder)
```

This shows it contains economic and demographic information about different countries across years.  There are 1704 rows (observations) and 6 columns (variables).

Each variable name is listed along with a variable type designation.  

* fct: means a factor variable, also known as a categorical variable.
* int: means a quantitative variable that takes only integer or whole number values.
* dbl: means double precision, a quantitative variable that is essentially continuous - taking decimal values.

## head 

By default, the `head` command will show the first 6 rows of the dataset gapminder. Datasets in R are called "dataframes."  The gapminder dataframe is denoted as a "tibble" which is a type of dataframe.   

Options to the head command can change the rows displayed.
```{r}
# default is to show 6 rows
head(gapminder)
# show only 4 rows...
head(gapminder,n=4)
```

## summary  

This command shows a basic summary of the values in each variable.

```{r}
# A basic, base R command
summary(gapminder)
```

The next command illustrates a "pipe" - here the dataframe gapminder is "piped" into
the summary function to be processed.  Note the same output is produce as using
`summary(gapminder)`.  Note, the pipe operation %>% is contained in tidyverse package: *magrittr* which is loaded when *tidyverse* is loaded.
```{r}
# Same idea, but using tidyverse pipe
gapminder %>% summary()
```


## Dataframe Details:  funModeling package

The *funModeling* package contains the `df_status` command which also summarizes
a dataframe - showing different aspects like missing values, percentage of zero 
values, and also the number of unique values.
```{r}
funModeling::df_status(gapminder)
```

```{r}
di=funModeling::data_integrity(gapminder)
# returns a detailed summary of all variables
print(di)
```


## Dataframe Details: skimr package

The *skimr* package contains many useful functions for summarizing a dataframe.  When we supply a dataframe to the `skim_without_charts` function, dataframe details are separated by variable types.

```{r, dfdetskimr1, eval=TRUE,render = knitr::normal_print}
gapminder %>% 
  skimr::skim_without_charts() 
```

## describe: Hmisc package

The *Hmisc* package contains the `describe` function that gives a helpful overview of numeric and categorical variables.

```{r, dfdethmisc}
gapminder %>% 
  Hmisc::describe() 
```

