# Univariate Graphical Displays {#UnivariateGraphicalDisplays}

```{r,echo=FALSE,message=FALSE,warning=FALSE}
library(gapminder)
library(tidyverse)
```


In this section we will show examples of how to create graphical displays of a single variable - with examples for both quantitative and categorical variables. 
In each example, the first line creates the dataset to be graphed - followed by a command making the display.  We will focus on graphical displays made by functions in the *ggplot2* family - that is, the *ggplot2* package which is also part of the *tidyverse* family of functions.  If *tidyverse* is loaded, *ggplot2* functions will work without explicitly loading the *ggplot2* package.

## Overview of ggplot

The *ggplot2* package uses the `ggplot` command - and builds a graphical display in steps and layers.  We always start with the `ggplot` command which typically has two basic elements:  a dataset to be used, and a list of mappings `aes` that is used to connect dataset variables to aspects of the plot like the vertical axis, horizontal axis, or perhaps the size of a point.

The kind of object being displayed is called a `geom`, and a plot can have several `geom`s, and they are added to a display in layers - connected by a `+` sign.


## A Quantitative Variable

### Dotplot

The next block of code takes the gapminder dataframe and "pipes" (%>%, a pipeline like plumbing) the data through a filter so that only data from year 1997 flows through to define the new dataset named `ds`.  The `ggplot` command uses dataset ds, and variable x life expectancy.  The next example shows what using only the `ggplot` command produces an empty graphical region that is awaiting further instructions:

```{r,message=FALSE,tidy=FALSE}
ds <- gapminder %>% filter(year==1997) 
#
ggplot(data=ds, mapping=aes(x=lifeExp)) 
```

Now we use additional code to place the dotplot in the existing graphical region.  In ggplot graphics we make graphical objects with a `geom` function - here a dotplot so we use `geom_dotplot()` to produce the dotplot specified using the variable mappings in the aesthetics command `aes` in the ggplot command.  

```{r,message=FALSE,tidy=FALSE}
ds <- gapminder %>% filter(year==1997) 
#
ggplot(data=ds, mapping=aes(x=lifeExp)) + 
  geom_dotplot() + 
  xlab("Life Expectancy (years)") + ylab("Frequency")
```

Here we change the default size for the dots.
```{r,message=FALSE,tidy=FALSE}
ds <- gapminder %>% filter(year==1997) 
#
ggplot(data=ds, mapping=aes(x=lifeExp)) + 
  geom_dotplot(dotsize=0.70) + 
  xlab("Life Expectancy (years)") + ylab("Frequency")
```


#### Dotplot with observations identified and ordered

Here we produce a display so that life expectancy is displayed for each country in Asia, and the values are ordered.
```{r}
ds <- gapminder %>% filter(continent=="Asia",year==1997) 
#  
ggplot(data=ds, mapping=aes(x=lifeExp, y= reorder(country,lifeExp))) + 
  geom_point() + 
  xlab("Life Expectancy (years) in 1997") + 
  ylab("Asian Countries")
```

Notice that in the next example we simply pipe the modified dataset into the first argument of the `ggplot` command so that there is no need to save the modified dataset to make the display.  In the next block we pipe the modified dataset directly inside the `ggplot` command to automatically replace the first argument.
```{r}
gapminder %>% filter(continent=="Asia",year==1997) %>%
ggplot(data=., mapping=aes(x=lifeExp, y= reorder(country,lifeExp))) + 
  geom_point() + 
  xlab("Life Expectancy (years) in 1997") + 
  ylab("Asian Countries")
```


### Histogram

This code block is similar to the dotplot commands, but the geom_histogram function controls the bin width in units of the x variable - in this case 5 years.
```{r,eval=TRUE,tidy=FALSE}
gapminder %>% filter(year==1997) %>%
ggplot(data=.,mapping=aes(x=lifeExp)) + 
  geom_histogram(binwidth=5) + 
  xlab("Life Expectancy (years)") +
  ylab("Relative Frequency")
```

Here we change the binwidth:
```{r,eval=TRUE,tidy=FALSE}
gapminder %>% filter(year==1997) %>%
ggplot(data=.,mapping=aes(x=lifeExp)) + 
  geom_histogram(binwidth=2.5) + 
  xlab("Life Expectancy (years)") +
  ylab("Relative Frequency")
```

Here we change the number of bins, notice the data argument is replaced by the piped data:
```{r,eval=TRUE,tidy=FALSE}
gapminder %>% filter(year==1997) %>%
ggplot(mapping=aes(x=lifeExp)) + 
  geom_histogram(bins=12) + 
  xlab("Life Expectancy (years)") +
  ylab("Relative Frequency")
```


### Density Plot

Density plots produces a smoothing of a histogram to display the distribution.
```{r,eval=TRUE,tidy=FALSE}
ds <- gapminder %>% filter(year==1997)
#
ggplot(data=ds, mapping=aes(x=lifeExp)) + 
  geom_density() + 
  xlab("Life Expectancy (years)") +
  ylab("Density")
```

The `adjust` option controls the amount of smoothing relative to a default value of 1.  A smaller value gives less smoothing (more responsive line to small changes in the data distribution), and larger values will make a smoother curve that is less sensitive to the data pattern.
```{r,eval=TRUE,tidy=FALSE}
ds <- gapminder %>% filter(year==1997)
#
ggplot(data=ds, mapping=aes(x=lifeExp)) + 
  geom_density(adjust=0.75) + 
  xlab("Life Expectancy (years)") +
  ylab("Density")
```

### Boxplot

The boxplot display really needs only a single quantitative variable (here life expectancy) for the numeric axis.  However, the other axis looks better with some 
sort of factor variable - so here we supply the year for the display, where the quantitative variable `year` has temporarily being used as a category/factor variable by being processed by the `factor` function before used in the graphic:
```{r,tidy=FALSE}
ds <- gapminder %>% filter(year==1997) 
#
ggplot(data=ds, mapping=aes(x=factor(year),y=lifeExp)) +
 geom_boxplot() + 
  labs(x="Year",y="Life Expectancy (years)")
# Change orientation
ggplot(data=ds, mapping=aes(x=factor(year),y=lifeExp)) +
 geom_boxplot() + 
  coord_flip() + 
  labs(x="Year",y="Life Expectancy (years)")
```

Now we overlay points on top of the boxplot display.  Note the `geom_jitter` that overlays the points has an argument `alpha=0.5` signifying a slightly transparent plot symbol.  An alpha value of 1 means the plot symbol is opaque, and a value of 0 is completely transparent.  Careful use of alpha in large datasets will enable the analyst to correctly perceive point density.  Without using a smaller value of `alpha` the plot may be one large blob of ink - making it difficult to judge the density of points in the display.

```{r,tidy=FALSE}
ds <- gapminder %>% filter(year==1997) 
#
ggplot(data=ds, mapping=aes(x=factor(year),y=lifeExp)) +
 geom_boxplot(outlier.shape = NA) + 
 geom_jitter(alpha=0.5, width=0.35) +
  labs(x="Year",y="Life Expectancy (years)")
```


If the dataframe has only one quantitative variable, we can make a character variable called "sample", then this code will produce an acceptable display.

```{r,tidy=FALSE}
ds <- gapminder %>% filter(year==1997) %>%
  mutate(sample="Sample")
#
ggplot(data=ds, mapping=aes(x=sample,y=lifeExp)) +
 geom_boxplot(outlier.shape = NA) + 
 geom_jitter(alpha=0.5, width=0.35) +
  labs(x="",y="Life Expectancy (years)")
```

## Displays of a Categorical Variable

### Bar Graph

```{r, tidy=FALSE}
ds <- gapminder %>% 
  filter(year==1997) %>% 
  group_by(continent) 
# Frequency of countries in each continent in 1997.
ggplot(data=ds, mapping=aes(x=continent)) + 
  geom_bar() +
  labs(x="Continent", y="Frequency")
#
ggplot(data=ds, mapping=aes(x=continent)) + 
  geom_bar(width=0.5,fill="blue") +
  labs(title="Countries in Each Continent",
       subtitle = "Year = 1997",
       caption = "Gapminder data",
       x="Continent", 
       y="Frequency") +
  theme(axis.text.x = element_text(angle=45,vjust = 0.6))
```

Bar graph with percentages on vertical axis.

```{r,tidy=FALSE,message=FALSE,warning=FALSE}
ds <- gapminder %>% 
  filter(year==1997) %>% 
  group_by(continent)  %>%
  summarise (n = n()) %>%
  mutate(pct = 100*n / sum(n)) 
#
head(ds)
#
ggplot(data=ds, mapping=aes(x = continent, y = pct)) + 
  geom_bar(stat = "identity") + 
  xlab("Continent") + ylab("Percentage")
# change order of continents in decreasing frequency order
ggplot(data=ds, mapping=aes(x = reorder(continent, -pct), y = pct)) + 
  geom_bar(stat = "identity") + 
  xlab("Continent") + ylab("Percentage")
```

Sometimes it is more convenient to have the bars oriented horizontally.  Notice we set up the aesthetic mappings as usual and then flip the axes with the `coord_flip` command.

```{r,tidy=FALSE,message=FALSE,warning=FALSE}
ds <- gapminder %>% 
  filter(year==1997) %>% 
  group_by(continent)  %>%
  summarise (n = n()) %>%
  mutate(pct = 100*n / sum(n)) 
#
ggplot(data=ds, mapping=aes(x = reorder(continent, pct), y = pct)) + 
  geom_bar(stat = "identity") +
  coord_flip() +
  xlab("Continent") + ylab("Percentage")
```

You can produce a similar display using `geom_col` on the summarized data tibble.
```{r}
ggplot(data = ds, mapping = aes(x = continent, y = n)) +
  geom_col() +
  ylab("Frequency")
```


### Pie Graph

Pie graphs are not recommended, but the code needed to make one is given here.
```{r,tidy=FALSE,message=FALSE,warning=FALSE}
contin.prop<- gapminder %>% 
  group_by(continent) %>%
  summarise (n = n()) %>%
  mutate(freq = n / sum(n))
#
ggplot(data=contin.prop, mapping=aes(x="",y=freq,fill=continent)) + 
  geom_bar(width=1,stat="identity") +
  coord_polar("y",start=0) +
  xlab("") + ylab("Country Frequency by Continent")
```

