# Summary Statistics For One Variable

```{r,echo=FALSE,message=FALSE,warning=FALSE}
library(gapminder)
library(tidyverse)
```

## One Quantitative Variable

### Using base R summary function

```{r,tidy=FALSE}
gapminder %>% filter(year==1997) %>% select(lifeExp) %>% summary()
```

### Using dplyr summarise function

It is often helpful to create data summaries during preliminary phases of examination.  Here is how to use the summarise command in the analysis pipeline system.

```{r,tidy=FALSE}
gapminder %>% filter(year==1997) %>% 
      dplyr::summarise(meanLE=mean(lifeExp,na.rm=TRUE),
                       medLE=median(lifeExp,na.rm=TRUE),
                       sd=sd(lifeExp,na.rm=TRUE),
                       iqr=IQR(lifeExp,na.rm=TRUE),
                      Q1=quantile(lifeExp,probs=0.25,na.rm=TRUE),
                      Q3=quantile(lifeExp,probs=0.75),
                      n=n())
```

### Summary Statistics Using funModeling package

The `profiling_num` and `plot_num` functions from the *funModeling* package help
give a concise numeric and visual overview of the numeric variables in the
dataframe.
```{r}
funModeling::profiling_num(gapminder)
funModeling::plot_num(gapminder)
```

This example shows summary statistics for two quantitative variables.  For only one variable, simply use `select` for only one variable.
```{r,tidy=FALSE}
gapminder %>%
  filter(year==1997) %>%
  select(lifeExp,gdpPercap) %>%
funModeling::profiling_num() 
#
gapminder %>%
  filter(year==1997) %>%
  select(lifeExp,gdpPercap) %>%
funModeling::plot_num() 
```





### Summary Statistics: skimr package

The *skimr* package produces summary statistics about variables and overviews for dataframes.  It is easy to manipulate and use pipes, select, and filter from the tidyverse family of packages.

The next code supplies a dataframe that contains both categorical variables (continent), and numeric variables (lifeExp, gdpPercap).  Numeric variables are chosen with the `yank` function, then some attributes are omitted from the display (n_missing, complete_rate) using the `select` function from dplyr.

```{r}
varlist <- c("n_missing","complete_rate")
gapminder %>% filter(year==1997) %>% 
  select(-year, -country, -pop) %>% 
  skimr::skim_without_charts() %>%
  skimr::yank("numeric") %>%
  dplyr::select(-one_of(varlist))
  
```




## One Categorical Variable

### Counting Values

The next command counts the number of rows in the dataset for each continent - then we show a variant which pipes the output into the kable function for a more attractive table.

```{r}
gapminder %>% count(continent)
#
gapminder %>%  count(continent) %>% knitr::kable()
#
gapminder %>%  count(continent, sort=TRUE) %>% knitr::kable()
```

The previous code tells us how many lines (rows) for each continent, but many rows are repeated for each country - just different years.

```{r}
gapminder %>% filter(year==1997 | year==1967) %>%
  dplyr::group_by(continent) %>%
  dplyr::summarise(n = n(), n_countries = n_distinct(country)) %>% knitr::kable()
```


### Categorical variable:  skimr package

Here we summarize a categorical variable (continent), and observe it has 5 unique values (levels) and the most frequent values are displayed.
```{r}
gapminder %>% filter(year==1997) %>% 
  select(lifeExp,continent) %>% 
  skimr::skim_without_charts() %>%
  skimr::yank("factor") %>%
  dplyr::select(-n_missing,-ordered,-complete_rate)
```

### Categorical variable: funModeling package

The *funModeling* package gives an easy way to learn about categorical variables of types:  character and factor.  There are two categorical variables in the gapminder
dataframe: country and continent.  There are a lot of countries, so we demonstrate
this command for only the continent variable.
```{r,warning=FALSE}
# Frequency distribution of entire dataframe
# will produce lots of output and warnings
#funModeling::freq(gapminder)
# next command for one category variable:  continent
funModeling::freq(gapminder$continent)
```

There are a lot of observations (rows) for Africa and very few for Oceania (Australia, New Zealand, etc).

### Categorical variable: janitor package

Let's begin with the base R function `table`:
```{r}
gapminder %>% 
  filter(year==1997) %>%
  select(continent) %>%
  table()
```

Now contrast with the `tabyl` function from the *janitor* package:

```{r}
gapminder %>% 
  filter(year==1997) %>%
  janitor::tabyl(continent,sort=TRUE) %>%
  knitr::kable()
#
gapminder %>% 
  filter(year==1997) %>%
  janitor::tabyl(continent,sort=TRUE) %>%
  janitor::adorn_pct_formatting(digits=2,affix_sign = TRUE) %>%
  knitr::kable()
```
