# Exploratory Data Analysis For One Quantitative Variable: by Groups

It is often helpful to create data summaries of a quantitative variable for each level of a grouping variable.

```{r,echo=FALSE,message=FALSE,warning=FALSE}
library(gapminder)
library(tidyverse)
```

## Summary Statistics: dplyr

Using *dplyr* and *tidyverse* for summary statistics across the levels of a group variable (of type factor/categorical) requires the use of the verb `group_by`.  Here we produce summary statistics of life expectancy across the levels of continent.

```{r,tidy=FALSE}
# Output presented in initial continent order (alphabetic)
gapminder %>% filter(year==1997) %>% 
  filter(continent != "Oceania") %>% 
  group_by(continent) %>% 
  summarise(meanLE=mean(lifeExp,na.rm=TRUE),
            medLE=median(lifeExp,na.rm=TRUE),
            sd=sd(lifeExp,na.rm=TRUE),
            iqr=IQR(lifeExp,na.rm=TRUE),
            Q1=quantile(lifeExp, probs=0.25,na.rm=TRUE),
            Q3=quantile(lifeExp,probs=0.75),
            n=n())
#
# Output rows ordered by decreasing values of a statistic (mean Life Expectancy):
gapminder %>% filter(year==1997) %>% 
  filter(continent != "Oceania") %>% 
  group_by(continent) %>%
  summarise(meanLE=mean(lifeExp,na.rm=TRUE),
            medLE=median(lifeExp,na.rm=TRUE),
            sd=sd(lifeExp,na.rm=TRUE),
            iqr=IQR(lifeExp,na.rm=TRUE),
            min=min(lifeExp),
            max=max(lifeExp),
            n=n())  %>%
  arrange(desc(meanLE))
```


Next, we save the statistics table to an object called statstable, then we use the `kable` function for display.

```{r,tidy=FALSE}
statstable <- gapminder %>% filter(year==1997) %>% 
  filter(continent != "Oceania") %>% 
  group_by(continent) %>% 
  summarise(meanLE=mean(lifeExp,na.rm=TRUE),
            medLE=median(lifeExp,na.rm=TRUE),
            sd=sd(lifeExp,na.rm=TRUE),
            iqr=IQR(lifeExp,na.rm=TRUE),
            min=min(lifeExp),
            max=max(lifeExp),
            n=n())  %>% 
  arrange(desc(meanLE))
#
knitr::kable(statstable)
```


## Summary Statistics:  skimr

Here we implement the `group_by` function to display descriptive statistics for numeric variables by continent, for two quantitative variables using functions from the *skimr* package.
```{r}
gapminder %>% filter(year==1997) %>% 
  filter(continent != "Oceania") %>% 
  group_by(continent) %>% 
  skimr::skim_without_charts() %>%
  skimr::yank("numeric") %>%
  dplyr::filter(skim_variable %in% c("lifeExp","gdpPercap")) %>%
  knitr::kable()
```

## Graphical Displays of a quantitative variable, separated by groups

In each example, the first lines create the dataset to be graphed - followed by a `ggplot` command making the display.  Several of the examples make use of the principle of "small-multiples" so that each level of the factor variable has a separarate panel for the quantitative variable display.

### Dotplots

```{r,message=FALSE,tidy=FALSE}
ds <- gapminder %>% filter(year==1997) 
#
ggplot(data=ds,mapping=aes(x=lifeExp)) + 
  geom_dotplot() + 
  facet_wrap( ~ continent,ncol=2) + 
  xlab("Life Expectancy (years)") +
  ylab("Frequency")
```


### Histograms

```{r,eval=TRUE,tidy= FALSE}
ds <- gapminder %>% 
  filter(year==1997) %>% 
  filter(continent != "Oceania") %>% 
  group_by(continent) 
#
ggplot(data=ds, mapping=aes(x=lifeExp)) + 
  geom_histogram(binwidth=5) + 
  facet_wrap( ~ continent,ncol=2) + 
  xlab("Life Expectancy (years)") +
  ylab("Frequency")
```


### Density Plots in Facets

The code given here shows how to produce a density plot in separate panels for each continent.

```{r,eval=TRUE,tidy=FALSE}
ds <- gapminder %>% 
  filter(year==1997) %>% 
  filter(continent != "Oceania") %>% 
  group_by(continent) 
#
ggplot(data=ds, mapping=aes(x=lifeExp, colour=continent, fill=continent)) + 
  geom_density(alpha = 0.35) + 
  xlab("Life Expectancy (years)") +
  ylab("Density") +
  facet_wrap( ~ continent, ncol = 2) +  
  theme(legend.position = "none")
```

### Overlaid Density Plots

The initial command below takes the gapminder data and consider only observations (rows) from 1997, but exclude all observations from Oceania.  The `alpha` setting controls the amount of transparency in the densities for each continent - smaller values of `alpha` (between 0 and 1) are more transparent.
```{r,eval=TRUE,tidy=FALSE}
gapminder %>% 
  filter(year==1997) %>% 
  filter(continent != "Oceania") %>% 
  group_by(continent) %>%
ggplot(mapping=aes(x=lifeExp, colour=continent, fill=continent)) + 
  geom_density(alpha = 0.35) + 
  xlab("Life Expectancy (years)") +
  ylab("Density")
```


### Boxplots, Grouped Data

In the code below, the `alpha` value again controls the transparency of the points alpha=1 means opaque, alpha=0 means completely see-through.  When there is a lot of data, use a smaller value of alpha.

```{r,tidy=FALSE}
ds <- gapminder %>% 
  filter(year==1997) %>% 
  filter(continent != "Oceania") %>% 
  group_by(continent) 
#
ggplot(data=ds, mapping=aes(x=continent,y=lifeExp)) +
 geom_boxplot() + 
  labs(x="Continent",y="Life Expectancy (years)")
#
ggplot(data=ds, mapping=aes(x=continent,y=lifeExp)) +  
 geom_boxplot(outlier.colour = NA) + 
 geom_point(position = position_jitter(width = 0.15, height = 0.15),alpha=.50) +
  labs(x="Continent",y="Life Expectancy (years)")
```

### Boxplots, overlay points on the boxplots with color control

In the code below, the alpha value controls the transparency of the points alpha=1 means opaque, alpha=0 means completely see-through. 

```{r oztempbox2, tidy=FALSE}
ds <- gapminder %>% 
  filter(year==1997) %>% 
  filter(continent != "Oceania") %>% 
  group_by(continent) 
# 
ggplot(data=ds, mapping=aes(x=continent,y=lifeExp, colour=continent)) + 
  geom_point(position = position_jitter(width = 0.2, height = 0.2),alpha=.25) +
 geom_boxplot(outlier.colour = NA, fill = NA) + 
  labs(x="Continent",y="Life Expectancy (years)")
# 
ggplot(data=ds, mapping=aes(x=continent,y=lifeExp, colour=continent)) + 
  geom_point(position = position_jitter(width = 0.2, height = 0.2),alpha=.80) +
 geom_boxplot(outlier.colour = NA, fill = NA) + 
  labs(x="Continent",y="Life Expectancy (years)")
```
