# Analysis of One Categorical Variable by another categorical variable

```{r,echo=FALSE,message=FALSE,warning=FALSE}
library(gapminder)
library(tidyverse)
library(fivethirtyeight)
```

To demonstrate graphical displays of two categorical variables, we need a new dataset with two categorical variables.  We use the `congress_age` dataframe from the *fivethirtyeight* package.  In these displays we will use categorical variables:

* party affiliation (party) with values: D, I, R.
* congressional chamber (chamber) with values: house, senate

We will restrict ourselves to the 113th congress, a meeting of the legislative branch of the United States federal government, from January 3, 2013, to January 3, 2015, during the fifth and sixth years of Barack Obama's presidency.

## Tables

```{r}
congage <- fivethirtyeight::congress_age
ds1 <- congage %>% filter(congress > 112) %>% select(congress,chamber,state,party,incumbent,age) 
#  We declare party and chamber as factor/categorical variables, and control their levels.
ds1 <- ds1 %>% mutate(party = factor(party,levels=c("D","I","R")),
                    chamber = factor(chamber))
ds1 <- ds1 %>% na.omit()
ds <- ds1  
#
table(ds$chamber,ds$party)
#
mytable <- table(ds$chamber,ds$party)
#
prop.table(mytable) # cell percentages
prop.table(mytable, 1) # row percentages
prop.table(mytable, 2) # column percentages
```

```{r}
ds %>%  janitor::tabyl(chamber, party)
#
t2 <- ds %>%  janitor::tabyl(chamber, party)
t2 %>%
  janitor::adorn_percentages("row") %>%
  janitor::adorn_pct_formatting(digits = 2) %>%
  janitor::adorn_ns()
# column percentages
t2 %>%
  janitor::adorn_percentages("col") %>%
  janitor::adorn_pct_formatting(digits = 2) %>%
  janitor::adorn_ns()
# both row and column percentages
t2 %>%
  janitor::adorn_percentages("all") %>%
  janitor::adorn_pct_formatting(digits = 2) %>%
  janitor::adorn_ns()
```

```{r,tidy=FALSE}
congage <- fivethirtyeight::congress_age
ds1 <- congage %>% filter(congress > 112) %>% select(congress,chamber,state,party,incumbent,age) 
#  We declare party and chamber as factor/categorical variables, and control their levels.
ds1 <- ds1 %>% mutate(party = factor(party,levels=c("D","I","R")),
                    chamber = factor(chamber))
ds1 <- ds1 %>% na.omit()
ds <- ds1  
ds %>% group_by(chamber,party) %>% 
  dplyr::count() %>% 
  tidyr::pivot_wider(names_from = party, values_from = n)
```


## Graphical Displays

```{r,tidy=FALSE}
# basic bar plot of party affiliation
ggplot(data=ds, aes(x=party)) + 
  geom_bar() +
  labs(x="Party", y="Count")
```


```{r, tidy=FALSE}
ds <- ds1 %>% group_by(party,chamber)
#
ggplot(data=ds, aes(x=chamber)) + 
  geom_bar(aes(fill=party),position="dodge") +
  labs(x="Chamber", y="Count")
#
ggplot(data=ds, aes(x=party)) + 
  geom_bar(aes(fill=party)) +
  facet_wrap( ~ chamber) +
  labs(x="Party", y="Count")
```



```{r,tidy=FALSE,message=FALSE,warning=FALSE}
#  The next display attempts to use percentages on the vertical axis defined within chamber.
# This means the next command must list chamber as the FIRST group_by variable.
ds <- ds1 %>% group_by(chamber,party)  %>%
   summarise (n = n()) %>%
  mutate(pct = 100*n / sum(n)) 
#
ds
#
ggplot(data=ds, aes(x=party, y=pct)) + 
  geom_bar(aes(fill=party),stat="identity") +
  facet_wrap( ~ chamber) +
  labs(x="Party", y="Percent")
```




```{r,tidy=FALSE,message=FALSE,warning=FALSE}
ds1 <- congage %>% filter(congress > 112) %>% select(congress,chamber,state,party,incumbent,age) %>%
 mutate(party = factor(party,levels=c("D","I","R")),
                    chamber = factor(chamber)) %>% 
  na.omit()
#
ds <- ds1 %>% group_by(chamber,party)  %>%
   summarise (n = n()) %>%
  mutate(pct = 100*n / sum(n)) 
#
ggplot(data=ds, mapping=aes(x=chamber,y=pct,fill=party)) + 
  geom_col() +
  labs(x="Chamber", y="Percent")
```

```{r,tidy=FALSE,message=FALSE,warning=FALSE}
ds1 <- congage %>% filter(congress > 112) %>% select(congress,chamber,state,party,incumbent,age) %>%
 mutate(party = factor(party,levels=c("D","I","R")),
                    chamber = factor(chamber)) %>% 
  na.omit()
#
ds1 %>% group_by(chamber,party)  %>%
   summarise (n = n()) %>%
  mutate(pct = 100*n / sum(n)) %>%
ggplot( mapping=aes(x=chamber,y=pct,fill=party)) + 
  geom_col() +
  coord_flip() +
  labs(x="Chamber", y="Percent")
```



