--- 
title: "Essential R Skills"
author: "Jon Anderson"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: jonander99/R-Essentials
description: "This book provides R code examples expected in all UMN-Morris Statistics courses."
---

# Motivation 

We have found that students enter our courses with wide variation in experience and comfort using statistical software for computation and making graphical displays.  This document represents our expectations for the basic R skills that students should know upon completing an introductory course in statistics.  Analysis methods may appear at times in this document, but the emphasis here is upon basic R usage for data wrangling, and exploratory data analysis using numeric and graphical methods.


```{r eval=FALSE,echo=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'tidyverse', 'gapminder'
), 'packages.bib')
```
